\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}

### 2012-04-06 pkg/evtree/Submission_JSS/Revisions/response_revision_14032012.tex

\documentclass[DIN, pagenumber=false, parskip=half,% 
               fromalign=left, fromphone=true,%  
               fromemail=true, fromurl=false, %
               fromlogo=true, fromrule=false, fromrule=afteraddress]{scrlttr2}
\usepackage[latin1]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}

\setkomavar{fromname}{Thomas Grubinger\\ Department of Medical Statistics, Informatics and Health Economics \\ Innsbruck Medical University}
\setkomavar{fromaddress}{Sch\"opfstrasse 41/1, 6020 Innsbruck, Austria}
\setkomavar{backaddress}{Dept.~of Med.~Statistics \\ Innsbruck Medical University\\ Sch\"opfstr. 41/1, A-6020 Innsbruck}
\setkomavar{fromphone}{+43 512 9003 70906}
\setkomavar{fromemail}{Thomas.Grubinger@i-med.ac.at}
\setkomavar{signature}{Thomas Grubinger, Achim Zeileis, Karl-Peter Pfeiffer}
\setkomavar{fromlogo}{ \includegraphics*[width=2.8cm]{logo}}
\setkomavar{subject}[]{Revision of  JSS 817}

@@

\begin{document}

\begin{letter}{
  Jan de Leeuw\\
  Department of Statistics\\
  University of California, Los Angeles\\
  United States of America
}

\opening{Dear Prof.~de Leeuw,}

please find attached the revised version of our manuscript
``evtree: Evolutionary Learning of Globally Optimal Classification and Regression Trees in R''
(JSS 817). We would like to thank the reviewer for the careful review and the constructive
comments. We have incorporated all of the reviewer's recommendations, in particular by
cleaning up the notation and providing some more discussion of the illustrations. We
hope to have addressed all concerns in our revision and provide a detailed
point-to-point reply below.

Sincerely,\\[-2cm]
\closing{}

\newpage

\textbf{\large Point-to-Point Reply}

\bigskip

\textbf{1. Overall Thoughts}

@@

\textbf{1. Overall Thoughts}

\textit{The paper did a good job of introducing a new tree-based algorithm
overall. A lot of details pertaining to traditional tree models and the concept
of evolutionary algorithms were covered, some of which could be clarified
further. A clean-up in notation and wording would be greatly beneficial. The
code appears easy-to-read and uses parameters consistent with other tree
packages.}

Thanks for pointing these issues out. In fact, providing a unifying but still
simple and easily accessible notation for presenting the framework was one
important objective of the manuscript. While preparing the manuscript, we had
revised the notation a couple of times and apparently did not clean it up
carefully enough. Hence, we appreciate the thorough review and followed
all of its recommendations in this revision. Details of the revision are
provided below.

\medskip

\textit{In the end, the authors gave honest comparisons of their methods to the
traditional rpart and ctree methods, but it is unclear as to when and why evtree
outperforms its competitors apart from the specific chessboard example given in
Section~5.2.}

As the forward-search algorithms just optimze over a subset of conceivable
trees in each step compared to global-search algorithms, the latter are in principle
able to outperform the former when the optimal solution contains more complex
interaction patterns with low main effects. This is now discussed explicitly
in the following new paragraph in Section~5.1:

``From these results, it is not obvious which characteristics drive \texttt{evtree}'s
relative performance. Presumably, for some datasets the forward-search algorithms already
yield trees that are close to optimal, thus leaving little room for further
improvements. In contrast, for other datasets with more complex interaction patterns
(and possibly low main effects) \texttt{evtree}'s global-search strategy is probably
able to provide better predictive accuracy and/or sparser trees.''

\bigskip

\textbf{2. Spelling/Grammatical Errors}

Thanks for catching these. We have corrected all of them and ran again a spell
checker on the manuscript.

\bigskip

\textbf{3. Notation}

@@

\textit{In Section 2, the authors introduce some notation for the tree setup,
but the notation seems inconsistent in the paper. The authors write
$\theta=(v_{n_1}, s_{n_1},\dots,v_{n_{M-1}},s_{n_{M-1}})$, where $n_r \in
\{1,\dots, M_{max}-1\}$ but then also write $v_r$ and $s_r$ where
$r=1,\dots,M-1$. Further, in Section 2.2, the authors talk about $s_r$ and $v_r$
where $r \in \{n_1, \dots, n_{M-1} \}$, which is immediately different from the
indexing given in equation (3). The authors should use a consistent notation for
indexing the $v$ and $s$.}

We cleaned up the notation and a binary tree is now consistently denoted
as $\theta=(v_1, s_1, \dots, v_{M-1}, s_{M-1})$, 
where $v_r \in \{1, \dots ,P\}$ are the splitting \emph{variables} and $s_r$ the
associated \emph{split} rules for the internal \emph{nodes} $r \in \{1,...,M-1\}$.
Thus, in order to keep the notation as simple as possible, the internal nodes of the
tree are just labeled with their number $r \in \{1,...,M-1\}$.

\bigskip

\textbf{4. Case Study}

\textit{The authors have provided very minimal information on the dataset.
(p.~14) The authors write ``Thus, the evolutionary tree uses a different cutoff
in art for book club members that joined in the last year as opposed to older
customers." The text does not explain which variable corresponds to ``join
date", but based on Figure 3, one can indirectly infer that this variable is
first. The authors have described te variable first as the number of months
since the first purchase, but do not say that members join the book club upon
making their first purchase.}

We added a table (Table 2), which describes the variables of the
Bookbinder's Book Club dataset. Also, we avoid the formulation ``join date''
in the text and use ``first purchase'' instead.

\medskip

@@

\textit{The table comparing the misclassification and evaluation function values
for the different methods identify the methods as evtree, rpart, ctree, rpart2,
and ctree2. The paragraph which describes this table instead identifies the
methods as ev, rp, ct, rp2, and ct2. It would be better to make sure the two are
consistent.}

We changed the identifiers in the describing paragraph to also use \texttt{evtree},
\texttt{rpart}, \texttt{ctree}, \texttt{rpart2}, and \texttt{ctree2} instead of the
object names.

\bigskip

\textbf{5. Miscellaneous}

Thanks for pointing all of these out. All of the reviewer's recommendations
have been incorporated into the revised version of the manuscript. The only
point that needs additional explanation is the following:

\medskip

\textit{(p.~10) Third paragraph: Ambiguous ``they" in ``However, as they
represent one of the best..., they are both...". Does ``they" refer to the
parent and offspring trees? Does this mean that despite the $(1+1)$ rule, both
the parent and offspring are both selected to the next generation? This seems
unclear.}

@@

Yes, ``they'' refer to the parent and the offspring. However, the sentence
explains the strong selective pressure of a $(\mu + \lambda)$ strategy with
$\mu>1$ and $\lambda>1$. With a $(1+1)$ strategy, the parent and the offspring
can never be selected for the next round (as it is stated in the last sentence
of the paragraph).

For better clarification, we replaced ``they'' with ``the parent and the
offspring" and changed the sentence:

``Using a $(\mu + \lambda)$ strategy, a strong selective pressure can occur in
situations as follows."

@@

with:

``Using a $(\mu + \lambda)$ strategy with $\mu>1$ and $\lambda>1$, a strong
selective pressure can occur in situations as follows.".

### 2012-04-06 pkg/evtree/inst/doc/evtree.Rnw
Some touch-ups, rephrasings, and modifications. In partiular, the letter contains some more introductory remarks/comments and streamlines some of the straightforward changes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@289 255f4249-2263-4417-b1ea-8498779e27a7

<<setup, echo=FALSE>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("rpart")
library("evtree")
data("BBBClub", package = "evtree")
cache <- FALSE
@

@@

To fill this gap, we introduce a new \proglang{R} package \pkg{evtree}, 
available from the Comprehensive \proglang{R} Archive Network at
\url{http://CRAN.R-project.org/package=evtree}, providing evolutionary
methods for learning globally optimal classification and regression trees.
Generally speaking, evolutionary algorithms are inspired by natural Darwinian
evolution employing concepts such as inheritance, mutation, and natural selection.
They are population-based, i.e., a whole collection of candidate solutions -- trees in this application -- is
processed simultaneously and iteratively modified by \textit{variation operators}
called \textit{mutation} (applied to single solutions) and \textit{crossover}
(merging different solutions). Finally, a survivor selection process favors
solutions that perform well according to some quality criterion, usually
called \textit{fitness function} or \textit{evaluation function}. In this
evolutionary process the mean quality of the population increases over time
\citep{baeck,eiben2003iec}.  
In the case of learning decision trees, this means that the variation
operators can be applied to modify the tree structure (e.g., number of splits, splitting
variables, and corresponding splitpoints etc.) in order to optimize a
fitness functions such as the misclassificaion or error rate penalized by
the complexity of the tree.
A notable difference to comparable algorithms is the survivor selection
mechanism where it is important to avoid premature convergence. In the
following, we use a simple $(1+1)$ selection strategy (i.e., one parent
solution competes with one offspring for a place in the population)
which can be argued to offer computational
advantages for the application to classification and regression trees.

@@

To show why forward-search recursive partitioning algorithms typically
lead to globablly suboptimal solutions, their parameter spaces and optimization
problems are presented and contrasted in a unified notation. Although all
arguments hold more generally, only binary tree models with some maximum number
of terminal nodes $M_{\max}$ are considered. Both restrictions make the notation somewhat
simpler while not really restricting the problem: (a)~Multiway splits are equivalent
to a sequence of binary splits in predictions and number of resulting subsamples;
(b)~the maximal size of the tree is always limited by the number of observations
in the learning sample.

In the following, a binary tree model with $M$ terminal nodes (which consequently has
$M-1$ internal splits) is denoted by
%
\begin{equation}
  \theta ~=~ (v_{1}, s_{1},...,v_{M-1},s_{M-1} ), \label{eq:tree}
\end{equation}
%
where $v_r \in \{1, \dots ,P\}$ are the splitting \emph{variables} and $s_r$ the
associated \emph{split} rules for the internal \emph{nodes} $r \in \{1,...,M-1\}$.
Depending on the domain of $X_{v_r}$, the split rule $s_r$ contains either a cutoff
(for ordered and numeric variables) or a non-empty subset of $\{1, \dots, c\}$
(for a categorical variable with $c$~levels), determining which observations are
sent to the first or second subsample. In the former case, there are $u-1$ possible
split rules if $X_{v_r}$ takes $u$ distinct values; and in the latter case, there
are $2^{c-1} - 1$ possible splits. Thus, the product of all of these combinations
forms all potential elements $\theta$ from $\Theta_M$, the space of conceivable
trees with $M$ terminal nodes. The overall parameter space is then 
$\Theta =\bigcup_{M=1}^{M_{\max}} \Theta_{M}$ (which in practice is often reduced
by exluding elements $\theta$ resulting in too small subsamples etc.).

@@

\setkeys{Gin}{width=0.5\textwidth}
\begin{figure}[t!]
\centering
<<chess22-plot, fig=TRUE, echo=FALSE, height=4, width=4>>=
plot(X2 ~ X1, data = chess22, xlim = c(0, 2), ylim = c(0, 2), pch = c(1, 4)[Y], col = c("black", "slategray")[Y])
@
\caption{\label{fig:illustration} Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black circles and gray crosses.}
%%\vspace*{-0.3cm} %%FIXME: to avoid \pagebreak
\end{figure}

@@

@@

The principal function of the \pkg{evtree} package is the eponymous function
\code{evtree()} taking arguments
%
\begin{Code}
  evtree(formula, data = list(), weights = NULL, subset = NULL,
    control = evtree.control(...), ...)
\end{Code}
%
where \code{formula}, \code{data}, \code{weights}, and \code{subset} specify
the data in the usual way, e.g., via \code{formula = y ~ x1 + x2}.
Additionally, \code{control} comprises a list of control parameters
%
\begin{Code}
  evtree.control(minbucket = 7L, minsplit = 20L, maxdepth = 9L, 
    niterations = 10000L, ntrees = 100L, alpha = 1,
    operatorprob = list(pmutatemajor = 0.2, pmutateminor = 0.2,
      pcrossover = 0.2, psplit = 0.2, pprune = 0.2),
    seed = NULL, ...)
\end{Code}
%
where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth}
constrain the solution to a minimum number of observations in each terminal
node, a minimum number of observations in each internal node, and a maximum tree
depth. Note that the memory requirements increase by the square of the maximum
tree depth. Parameter \code{alpha} regulates the complexity parameter
$\alpha$ in Equations~\ref{eq:classification} and~\ref{eq:regression}, respectively.
\code{niterations} and \code{ntrees} specify the maximum number of iterations
and the number of trees in the population, respectively. With the argument
\code{operatorprob}, user-specified probabilities for the variation operators
can be defined. For making computations reproducible, argument \code{seed} is
an optional integer seed for the random number generator (at \proglang{C++} level).
If not specified, the random number generator is initialized by \verb|as.integer(runif(1, max = 2^16))|
in order to inherit the state of \code{.Random.seed} (at \proglang{R} level).
If set to \code{-1L}, the seed is initializied by the system time.

@@

\begin{table}[t!]
\centering
\begin{tabular}{ll}
\toprule	
     \textbf{Variable} & Description\\
\midrule	
     \textbf{amount} & total amount of money spent at the BBB Club. \\
     \textbf{art} & number of art books purchased. \\
     \textbf{child} & number of children's books purchased. \\
     \textbf{cook} & number of cookbooks purchased. \\
     \textbf{diy} & number of do-it-yourself books purchased. \\    
     \textbf{first} & number of months since the first purchase. \\   
     \textbf{freq} & number of books purchased at the BBB Club. \\          
     \textbf{gender} & factor indicating gender. \\
     \textbf{last} & number of months since the last purchase. \\         
     \textbf{youth} & number of youth books purchased. \\
\bottomrule
\end{tabular}
\caption{\label{tab:BBBClubDescription} This table describes the predictor variables of the \textit{Bookbinder's Book Club} dataset.} 
\end{table}

@@

Not surprisingly, the explanatory variable \code{art} -- the number of art books purchased previously at the
book club -- plays a key role in all constructed classification trees along with the number
of months since the first purchase (\code{first}), the frequency of previous purchases
(\code{freq}), and the customer's \code{gender}. Interestingly, though, the forward-search trees select
the arguably most important variable in the first split while the evolutionary tree uses \code{first}
in the first split and \code{art} in both second splits.
Thus, the evolutionary tree uses a different cutoff in \code{art} for book club members that bought there
first book in the last year as opposed to older customers.
While the former are predicted to be buyers if they had previously bought
at least one art book, the latter are predicted to purchase the advertised art book only if they
had previously bought at least two other art books. Certainly, this classification is easy to understand
and communicate (helped by Figure~\ref{fig:example_evtree}) to practitioners.

@@

Figure~\ref{fig:benchmark} (lower panels) summarizes the relative performance of \code{evtree} and \code{ctree}. For 15 out of 17 datasets \code{evtree} shows a better predictive performance. The algorithms' performances  is significantly worse on the \textit{MEL0101} dataset, where the mean squarred error of \code{ctree} is $5.6\%$ lower. However, on this dataset, \code{ctree} models are on average $86.5\%$ larger than \code{evtree} models. The relative complexity of \code{evtree} models is significantly smaller for 15 and larger for 1 dataset.

@@

No obvious relationship between the data characteristics and \code{evtree}'s relative
performance could be observed. Probably, one explanation is that for some datasets,
the recursive partitioning algorithms already yield (nearly) optimal trees, leaving little room for further
improvements. For other datasets, \code{evtree}'s global search strategy can find more
complex interactions, resulting in better predictive accuracy and/or sparser trees.

@@

%%\pagebreak %%FIXME

>>>>

<<setup, echo=FALSE>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("rpart")
library("evtree")
data("BBBClub", package = "evtree")
cache <- TRUE
@

@@

To fill this gap, we introduce a new \proglang{R} package \pkg{evtree}, 
available from the Comprehensive \proglang{R} Archive Network at
\url{http://CRAN.R-project.org/package=evtree}, providing evolutionary
methods for learning globally optimal classification and regression trees.
Generally speaking, evolutionary algorithms are inspired by natural Darwinian
evolution employing concepts such as inheritance, mutation, and natural selection.
They are population-based, i.e., a whole collection of candidate solutions -- trees in this application -- is
processed simultaneously and iteratively modified by \textit{variation operators}
called \textit{mutation} (applied to single solutions) and \textit{crossover}
(merging different solutions). Finally, a survivor selection process favors
solutions that perform well according to some quality criterion, usually
called \textit{fitness function} or \textit{evaluation function}. In this
evolutionary process the mean quality of the population increases over time
\citep{baeck,eiben2003iec}.  
In the case of learning decision trees, this means that the variation
operators can be applied to modify the tree structure (e.g., number of splits, splitting
variables, and corresponding split points etc.) in order to optimize a
fitness functions such as the misclassificaion or error rate penalized by
the complexity of the tree.
A notable difference to comparable algorithms is the survivor selection
mechanism where it is important to avoid premature convergence. In the
following, we use a simple $(1+1)$ selection strategy (i.e., one parent
solution competes with one offspring for a place in the population)
which can be argued to offer computational
advantages for the application to classification and regression trees.

@@

To show why forward-search recursive partitioning algorithms typically
lead to globally suboptimal solutions, their parameter spaces and optimization
problems are presented and contrasted in a unified notation. Although all
arguments hold more generally, only binary tree models with some maximum number
of terminal nodes $M_{\max}$ are considered. Both restrictions make the notation somewhat
simpler while not really restricting the problem: (a)~Multiway splits are equivalent
to a sequence of binary splits in predictions and number of resulting subsamples;
(b)~the maximal size of the tree is always limited by the number of observations
in the learning sample.

In the following, a binary tree model with $M$ terminal nodes (which consequently has
$M-1$ internal splits) is denoted by
%
\begin{equation}
  \theta ~=~ (v_{1}, s_{1},...,v_{M-1},s_{M-1} ), \label{eq:tree}
\end{equation}
%
where $v_r \in \{1, \dots ,P\}$ are the splitting \emph{variables} and $s_r$ the
associated \emph{split} rules for the internal \emph{nodes} $r \in \{1,...,M-1\}$.
Depending on the domain of $X_{v_r}$, the split rule $s_r$ contains either a cutoff
(for ordered and numeric variables) or a non-empty subset of $\{1, \dots, c\}$
(for a categorical variable with $c$~levels), determining which observations are
sent to the first or second subsample. In the former case, there are $u-1$ possible
split rules if $X_{v_r}$ takes $u$ distinct values; and in the latter case, there
are $2^{c-1} - 1$ possible splits. Thus, the product of all of these combinations
forms all potential elements $\theta$ from $\Theta_M$, the space of conceivable
trees with $M$ terminal nodes. The overall parameter space is then 
$\Theta =\bigcup_{M=1}^{M_{\max}} \Theta_{M}$ (which in practice is often reduced
by excluding elements $\theta$ resulting in too small subsamples etc.).

@@

\setkeys{Gin}{width=0.5\textwidth}
\begin{figure}[t!]
\centering
<<chess22-plot, fig=TRUE, echo=FALSE, height=4, width=4>>=
plot(X2 ~ X1, data = chess22, xlim = c(0, 2), ylim = c(0, 2), pch = c(1, 4)[Y], col = c("black", "slategray")[Y])
@
\caption{\label{fig:illustration} Class distribution of  the $(X_1, X_2)$-plane. The two classes are indicated by black circles and gray crosses.}
\end{figure}

@@

\pagebreak %% FIXME

@@

The principal function of the \pkg{evtree} package is the eponymous function
\code{evtree()} taking arguments
%
\begin{Code}
  evtree(formula, data = list(), weights = NULL, subset = NULL,
    control = evtree.control(...), ...)
\end{Code}
%
where \code{formula}, \code{data}, \code{weights}, and \code{subset} specify
the data in the usual way, e.g., via \code{formula = y ~ x1 + x2}.
Additionally, \code{control} comprises a list of control parameters
%
\begin{Code}
  evtree.control(minbucket = 7L, minsplit = 20L, maxdepth = 9L, 
    niterations = 10000L, ntrees = 100L, alpha = 1,
    operatorprob = list(pmutatemajor = 0.2, pmutateminor = 0.2,
      pcrossover = 0.2, psplit = 0.2, pprune = 0.2),
    seed = NULL, ...)
\end{Code}
%
where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth}
constrain the solution to a minimum number of observations in each terminal
node, a minimum number of observations in each internal node, and a maximum tree
depth. Note that the memory requirements increase by the square of the maximum
tree depth. Parameter \code{alpha} regulates the complexity parameter
$\alpha$ in Equations~\ref{eq:classification} and~\ref{eq:regression}, respectively.
\code{niterations} and \code{ntrees} specify the maximum number of iterations
and the number of trees in the population, respectively. With the argument
\code{operatorprob}, user-specified probabilities for the variation operators
can be defined. For making computations reproducible, argument \code{seed} is
an optional integer seed for the random number generator (at \proglang{C++} level).
If not specified, the random number generator is initialized by \verb|as.integer(runif(1, max = 2^16))|
in order to inherit the state of \code{.Random.seed} (at \proglang{R} level).
If set to \code{-1L}, the seed is initialized by the system time.

@@

\begin{table}[t!]
\centering
\begin{tabular}{ll}
\toprule	
     Variable & Description\\
\midrule	
     \code{choice} & Did the customer buy the advertised book? \\
     \code{amount} & Total amount of money spent at the book Club. \\
     \code{art}    & Number of art books purchased. \\
     \code{child}  & Number of children's books purchased. \\
     \code{cook}   & Number of cookbooks purchased. \\
     \code{diy}    & Number of do-it-yourself books purchased. \\    
     \code{first}  & Number of months since the first purchase. \\   
     \code{freq}   & Number of books purchased at the book Club. \\          
     \code{gender} & Factor indicating gender. \\
     \code{last}   & Number of months since the last purchase. \\         
     \code{youth}  & Number of youth books purchased. \\
\bottomrule
\end{tabular}
\caption{\label{tab:BBBClubDescription} Variables of the Bookbinder's Book Club data.} 
\end{table}

@@

Not surprisingly, the explanatory variable \code{art} -- the number of art books purchased previously at the
book club -- plays a key role in all constructed classification trees along with the number
of months since the first purchase (\code{first}), the frequency of previous purchases
(\code{freq}), and the customer's \code{gender}. Interestingly, though, the forward-search trees select
the arguably most important variable in the first split while the evolutionary tree uses \code{first}
in the first split and \code{art} in both second splits.
Thus, the evolutionary tree uses a different cutoff in \code{art} for book club members that made their
first purchase during the last year as opposed to older customers.
While the former are predicted to be buyers if they had previously bought
at least one art book, the latter are predicted to purchase the advertised art book only if they
had previously bought at least two other art books. Certainly, this classification is easy to understand
and communicate (helped by Figure~\ref{fig:example_evtree}) to practitioners.

@@

Figure~\ref{fig:benchmark} (lower panels) summarizes the relative performance of \code{evtree} and \code{ctree}. For 15 out of 17 datasets \code{evtree} shows a better predictive performance. The algorithms' performances  is significantly worse on the \textit{MEL0101} dataset, where the mean squared error of \code{ctree} is $5.6\%$ lower. However, on this dataset, \code{ctree} models are on average $86.5\%$ larger than \code{evtree} models. The relative complexity of \code{evtree} models is significantly smaller for 15 and larger for 1 dataset.

@@

From these results, it is not obvious which characteristics drive \code{evtree}'s
relative performance. Presumably, for some datasets the forward-search algorithms already
yield trees that are close to optimal, thus leaving little room for further
improvements. In contrast, for other datasets with more complex interaction patterns
(and possibly low main effects) \code{evtree}'s global-search strategy is probably
able to provide better predictive accuracy and/or sparser trees.

@@
